Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-14 mr-review.Rmd
minor update

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

@@

Note that the uncertainty in the fitted values from 
the first regression is not considered when performing the second regression.
For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is 
incorrect in two-stage calculations. This observation led researchers to develop 
likelihood-based MR methods.

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020].

Limited information maximum likelihood model equations are shown below (Equations \@ref(eq:lm5) and \@ref(eq:lm6)).

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
  (\#eq:lm5)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
  (\#eq:lm6)
\end{equation}

They differ from the two-stage model because the two error terms, $\epsilon_i$ and $\tau_i$, have a bivariate normal
distribution.

CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. Model fitting then 
proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998].

We discuss PMR-Egger below in the context of methods that model horizontal pleiotropy.

@@

There are several ways that experimental data may violate one or more of the three MR assumptions.
Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}

>>>>

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
way to assess robustness of estimates in the presence of possible assumption violations.

The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

@@

Note that the uncertainty in the fitted values from 
the first regression is not considered when performing the second regression.
For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is 
incorrect in two-stage calculations. This and other observations led researchers to develop 
likelihood-based MR methods.

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925likelihood]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We discuss 
PMR-Egger below in the context of methods that account for horizontal pleiotropy. Immediately below, we 
discuss CoMM.

CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. In this model, the 
authors jointly model two linear mixed effects models, one relating the outcome to the SNPs and a second 
tying the exposure to the SNPs. Model fitting then 
proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998]. @yang_comm_2019, 
working with gene expression traits as exposures, found that CoMM better prioritized genetic variants 
relative to other methods [@yang_comm_2019].

@@

Now that we've discussed several MR methods, it is worthwhile to consider ways in which the standard MR assumptions may be violated. 
Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. Because W affecs Y through pathways that don't involve exposure X, this scenario violates the exclusion restriction assumption."}

knitr::include_graphics()
